MISUSES AND INACCURACIES OF PAGEL'S LAMBDA ML ESTIMATION METHODS

### 2020-04-23 Manuscript/Manuscript-2.Rmd
vsn 2

---
# Estimates of Phylogenetic Signal Based on Lambda are Often Inaccurate

@@

**Keywords**: Pagel's lambda, phylogenetic signal \hfill\break

**Short Title**: Inaccuracies in Pagel's Lambda \hfill\break

@@

For many macroevolutionary analyses, it is often of interest to quantify the degree to which phylogenetic signal is displayed in continuous traits. Several analytical tools have been developed for this purpose [e.g., @Gittleman1990; @Abouheif1999; @Pagel1999; @Blomberg_et_al2003; @Klingenberg2010; @Adams2014a], which differ primarily in how they characterize the phylogenetic dependency of trait variation among taxa. One commonly used statistical measure, *Kappa* ($K$), expresses the strength of phylogenetic signal as the ratio of observed trait variation to the trait variation conditioned on the phylogeny; which is then scaled by what is expected under Brownian motion given the phylogeny's size and shape [@Blomberg_et_al2003; also @Adams2014a]. Another approach, Pagel's $\lambda$ [@Pagel1999], uses maximum likelihood to fit the data to the phylogeny under some model of evolutionary change (typically Brownian motion). The inclusion of a scaling parameter, $\lambda$,  transforms the lengths of the internal branches of the phylogeny to improve the fit, and this parameter describes the degree of phylogenetic signal in the dataset [@Pagel1999; @Freckleton_et_al2002]. Pagel's $\lambda$ also has the appeal that it may be included when estimating the association of traits in a phylogenetic context, meaning that one may account for the degree of phylogenetic signal while conducting phylogenetic regression or ANOVA [see @Freckleton_et_al2002]. \hfill\break

Several studies have investigated the statistical properties of methods for estimating phylogenetic signal under various conditions [e.g., @Munkemuller_et_al2012; @Pavoine2012;  @DinizFilho2012; @MolinaVenegas2017; see also @Revell_et_al2008; @Revell2010]. These have largely focused on the ability of methods to detect the presence of phylogenetic signal (i.e., type I and type II error rates) under complex models of evolutionary change, across a range of phylogeny sizes, and with varying degrees of phylogenetic uncertainty or unresolved topologies. In terms of parameter estimation, Revell [-@Revell2010] found that regression parameters were accurately estimated when $\lambda$ was included during phylogenetic regression, and Munkmuller et al. [-@Munkemuller_et_al2012] demonstrated that estimates of phylogenetic signal obtained using various measures generally increased when input levels of phylogenetic signal were stronger. However, the precision of those estimates could not be determined, because the input levels of phylogenetic signal were simulated via a scaling factor ($w$) that was not directly comparable to the measures of phylogenetic signal being compared [see @Munkemuller_et_al2012]. Indeed, understanding the precision in estimating levels of phylogenetic signal is paramount, because biological inferences regarding the strength of phylogenetic signal in phenotypic datasets requires high precision of its estimation. One study [@Boettiger_et_al2012] found that estimates of Pagel's $\lambda$ displayed less variation when data were simulated on a large phylogeny ($N=281$) as compared to a small one ($N=13$), concluding that insufficient data (i.e., the number of species) was the underlying cause of the increased variation across parameter estimates. However, this conclusion assumes that the precision in estimating $\lambda$ remains constant across its range ($\lambda = 0 \to 1$); an assumption that to date, has not been verified. Thus, despite widespread use of Pagel's [-@Pagel1999] $\lambda$ in macroevolutionary studies, at present, we still lack a general understanding of the precision with which $\lambda$ can estimate levels of phylogenetic signal in phenotypic datasets. \hfill\break

In this study, we evaluate the precision of Pagel's $\lambda$ in estimating known levels of phylogenetic signal in phenotypic data. We use computer simulations with differing numbers of species, differently shaped phylogenies, and differing input levels of phylogenetic signal, to explore the degree to which $\lambda$ correctly identifies known levels of phylogenetic signal, and under what circumstances. We also evaluate the precision of a standardized effect size ($Z_K$) based on $Kappa$ for comparison. Additionally, we use simulations to determine how the inclusion of $\lambda$ in phylogenetic regression and ANOVA (i.e., PGLS) affects parameter estimation, and whether levels of phylogenetic signal estimated in a PGLS framework are accurate. We then survey the recent macroevolutionary literature for published papers containing estimates of $\lambda$ from empirical datasets, and compare these empirical estimates to patterns gleaned from our computer simulations. In general we find that while PGLS parameters (e.g., $\beta$) are accurately estimated with the inclusion of phylogenetic signal, estimates of $\lambda$ are not. We also find that estimates of $\lambda$ vary widely for a given input value of phylogenetic signal, and that this variation is not constant across the range of input signal, with decreased precision when phylogenetic signal is of intermediate strength. Additionally, the same $\lambda_{est}$ may be obtained from datasets containing vastly different input levels of phylogenetic signal. By contrast, variation across effect sizes ($Z_K$) obtained from $Kappa$ is far more consistent across the range of input values, and the precision of $Z_K$ is generally preferable to that of $\lambda_{est}$. Further, because $Z_K$ is a standardized effect size, the strength of phylogenetic signal is more readily interpretable, and may be compared across datasets. We conclude that estimates of phylogenetic signal using Pagel's $\lambda$ are often inaccurate, and thus interpreting strength of phylogenetic signal in phenotypic datasets based on this measure is compromised. By contrast, effect sizes obtained from $Kappa$ hold promise for characterizing phylogenetic signal, and for comparing the strength of phylogenetic signal across datasets.

>>>>

---
# A Standardized Effect Size for Evaluating the Strength of Phylogenetic Signal, and Why Lambda is not Appropriate

@@

**Keywords**: phylogenetic signal, effect size, Pagel's lambda \hfill\break

**Short Title**: An Effect Size for Phylogenetic Signal \hfill\break

@@

Several analytical tools have been developed to quantify phylogenetic signal in phenotypic datasets, including measures of serial independence [$\mathbf{C}$: @Abouheif1999], autocorrelation estimates [$I$: @Gittleman1990], statistical ratios of trait variation relative to what is expected given the phylogeny [$Kappa$: @Blomberg_et_al2003; @Adams2014a], and scaling parameters used in maximum likelihood fitting of the data to the phylogeny [$\lambda$: @Pagel1999], among others [e.g., @Klingenberg2010]. The statistical properties of these methods -- namely type I error rates and power -- have also been investigated to determine when phylogenetic signal can be detected and under what conditions [e.g., @Munkemuller_et_al2012; @Pavoine2012;  @DinizFilho2012; @Adams2014a; @MolinaVenegas2017; see also @Revell_et_al2008; @Revell2010]. One of the most widely used methods for characterizing phylogenetic signal in macroevolutionary studies is Pagel's $\lambda$ [@Pagel1999]. Here, maximum likelihood is used to fit the data to the phylogeny under a Brownian motion model of evolution. A parameter ($\lambda$) is included, which transforms the lengths of the internal branches of the phylogeny to improve the fit [@Pagel1999; @Freckleton_et_al2002]. Pagel's $\lambda$, which ranges from $0\to1$, is then treated as a measure of phylogenetic signal; with values close to 0 signifying little phylogenetic signal, and values close to 1 indicating that trait variation accumulates according to Brownian motion. Pagel's $\lambda$ also has the appeal that it may be included in phylogenetic regression  (PGLS) to account for the degree of phylogenetic signal in comparative analyses [see @Freckleton_et_al2002]. \hfill\break

Evolutionary biologists commonly seek to describe the relative strength of phylogenetic signal in phenotypic traits, to determine the extent to which shared evolutionary history has influenced trait covariation among taxa. This is often accomplished by interpreting empirical estimates of $\lambda$; with $\lambda\approx0$ signifying 'weak' phylogenetic signal, and $\lambda\approx1$ being interpreted as 'strong' phylogenetic signal [e.g., @DeMeester2019; @Pintanel2019;@Su2019]. Other approaches for interpreting $\lambda$ are more statistical, through the use of confidence intervals [@Vandelook2019] or likelihood ratio tests that compare the observed model fit to that obtained when $\lambda=0$ or $\lambda=1$ [@Freckleton_et_al2002; @Cooper2010; @Bose2019]. Likewise, qualitative comparisons of $\lambda$ across multiple phenotypic traits have also been used to infer whether the strength of phylogenetic signal is greater in one trait as compared to another [e.g., @Liu2019; @Bai2019]. Indeed, it seems intuitive to interpret the strength of phylogenetic signal in this manner, as $\lambda$ is a parameter on a bounded scale ($0\to1$) for which interpretation of its extremal points are well known ($\lambda=0$ is no phylogenetic signal, while $\lambda=1$ corresponds to Brownian motion). However, equating values of $\lambda$ directly to the strength of phylogenetic signal presumes two important statistical properties that have not been fully examined. \hfill\break

First, it presumes that values of $\lambda$ can be precisely estimated, and that parameter estimates of $\lambda$ are representative of the underlying phylogenetic signal. Understanding the precision in estimating levels of phylogenetic signal is paramount, because biological inferences regarding the strength of phylogenetic signal in phenotypic datasets requires high precision of its estimation. One study [@Boettiger_et_al2012] found that estimates of Pagel's $\lambda$ displayed less variation (i.e., greater precision) when data were simulated on a large phylogeny ($N=281$) as compared to a small one ($N=13$). From this observation it was concluded that insufficient data (i.e., the number of species) was the underlying cause of the increased variation across parameter estimates [@Boettiger_et_al2012]. Indeed, such a pattern is common with statistical estimators, as summary statistics and parameters are often more precise with greater sample sizes [@Cohen1988]. However, this conclusion also assumes that the precision in estimating $\lambda$ remains constant across its range ($\lambda = 0 \to 1$); an assumption that to date, has not been verified. Thus, despite widespread use of Pagel's [-@Pagel1999] $\lambda$ in macroevolutionary studies, at present, we still lack a general understanding of the precision with which $\lambda$ can estimate levels of phylogenetic signal in phenotypic datasets. \hfill\break

Second, while $\lambda$ is observed on a bounded scale ($0\to1$), this does not *de-facto* imply that the estimated values of this parameter correspond to the actual strength of the underlying input signal in the data. For this to be the case, $\lambda$ must be a statistical effect size. Effect sizes are a measure the magnitude of a statistical effect in data, represented on a common scale [@Glass1976; @Cohen1988]. Effect sizes have widespread use in many areas of the quantiative sciences, as they represent measures that may be readily summarized across datasets [e.g., meta-analysis: @HedgesOlkin1985; @Glass1976; @Arnqvist1995], or compared across datasets [e.g., @AdamsCollyer2016; @AdamsCollyer2019b]. Unfortunatley, not all statistical parameters and test statistics are effect sizes, and thus many summary measures must first be converted to standardized units (i.e., an effect size) for meaningful comparison [see @Rosenthal1994]. As a consequence, it follows that only if $\lambda$ is a statistical effect size can comparisons of estimates across datasets be interpretable. To date, it has not been determined whether $\lambda$ is an effect size for phylogenetic signal.  \hfill\break

In this study, we evaluate the precision of Pagel's $\lambda$ in estimating known levels of phylogenetic signal in phenotypic data. We use computer simulations with differing numbers of species, differently shaped phylogenies, and differing input levels of phylogenetic signal, to explore the degree to which $\lambda$ correctly identifies known levels of phylogenetic signal, and under what circumstances. We then derive a standardized effect size for phylogenetic signal, and apply the concept to two common measures of phylogenetic signal: $\lambda$ ($Z_{\lambda}$) and $Kappa$ ($Z_K$). In general we find that while PGLS parameters (e.g., $\beta$) are accurately estimated with the inclusion of phylogenetic signal, estimates of $\lambda$ are not. We also find that estimates of $\lambda$ vary widely for a given input value of phylogenetic signal, and that this variation is not constant across the range of input signal, with decreased precision when phylogenetic signal is of intermediate strength. Additionally, the same $\lambda_{est}$ may be obtained from datasets containing vastly different input levels of phylogenetic signal. Likewise, the precision of $Z_{\lambda}$ is less consistent than that observed with $Z_K$, across the range of input values, implying that $Z_K$ is a more robust effect size measure. Further, because $Z_K$ is a standardized effect size, the strength of phylogenetic signal is more readily interpretable, and may be compared across datasets. We conclude that estimates of phylogenetic signal using Pagel's $\lambda$ are often inaccurate, and thus interpreting strength of phylogenetic signal in phenotypic datasets based on this measure is compromised. By contrast, effect sizes obtained from $Kappa$ hold promise for characterizing phylogenetic signal, and for comparing the strength of phylogenetic signal across datasets.

# Methods and Results

-Lambda is not precise: holds in estimates, and PGLSregression
  -Precision varies with N (duh) but also WITH INPUT!
  -same input on different N, different lambda.est
  -same lambda.est from different input
  -Boettiger example: add small tips, greatly change lambda...
-Conclusion: Lambda does not quantify strength of phylogenetic signal as is often assumed

-Empirical examples: how interpretation can get folks ito trouble.

-A proposed Effect size for the strength of physig
  -Derivation of effect size (expand on Adams and Collyer 2019 suggestion that an effect size for phylogenetic signal is needed. The ‘gold standard’ of these is a standaredized effect size: (following the meta-analytic literature beginning with Glass 1976).  In other words, a Z-score. So, how does Z work when based on lambda? On kappa?
  - Z-score. could be Lambda or Kappa.  Show it is Kappa
-Comparing the strength of physig
  - two sample Z-score

-Conclusions and Implications

## $\lambda$ is not Precise

## Lambda does not measure phylogenetic signal

## An effect size for Physig
